File-based localization of numerical perturbations in data analysis pipelines

### 2020-09-23 paper.tex
Add the section of Availability of Supporting Data
Remove highlights of the revised sections

%%%% Commands %%%%
% \newcommand{\todo}[1]{\color{red}\textbf{TODO:}#1\color{black}}
% \newcommand{\note}[2]{\color{blue}Note from \textsc{#1}: #2\color{black}}
\newcommand{\revised}[1]{\color{blue}#1\color{black}\xspace}
\newcommand{\reprozip}[0]{ReproZip\xspace}
\newcommand{\tristan}[1]{\color{orange}\textbf{From Tristan:}#1\color{black}}
\newcommand{\toolname}[0]{Spot\xspace}
\newcommand{\flirt}[0]{\texttt{flirt}\xspace}
\newcommand{\fnirt}[0]{\texttt{fnirt}\xspace}

@@

\revised{Numerical perturbations resulting from variations in computational
environments} impact data analyses in various fields, but identifying the
origin of these perturbations in complex pipelines remains challenging.  In
some cases, small perturbations resulting from changes in operating system
versions~\cite{Glatard2015}, hardware~\cite{jezequel2015estimation}, or
parallelization parameters~\cite{diethelm2011limits}, result in
substantially different analysis outcomes, \revised{due to the propagation
and amplification of floating-point errors. While the existence of such
numerical errors is well known~\cite{stoer2013introduction}, their impact
on scientific computations has multiplied with the rise of the Big Data
era, due to the sustained growth of data sets, the increasing complexity of
analysis pipelines, and the diversification of computing infrastructures.}
To better understand and
correct these effects, efficient tools are
needed to assist pipeline developers
in the comparison of results obtained across different conditions.

@@

\begin{listing}
  \inputminted{bash}{"data/example/example.sh"}
  \caption{\revised{Example pipeline that computes the volume of the brain from a T1 image.}}
  \label{listing:sample-script}
\end{listing}

@@

We use \reprozip~\cite{rampin2016reprozip}
to capture: (1) the set of processes created by the
pipeline, and
(2) the set of files read and written by each process, including
temporary files. \reprozip collects this information through the
\texttt{ptrace()} system call, with no required instrumentation of the pipeline.
Using the \reprozip trace, \toolname reconstructs a provenance graph by creating process and file
nodes and by adding directed edges corresponding
to file reads and writes (Figure~\ref{fig:provenance-graph}).
\revised{We assume that provenance graphs are identical for the \reprozip traces obtained from
the same subjects in different operating systems.}

@@

\revised{
The labeled graph can differ depending on the order of executions in which condition we capture transient files 
or execute the pipeline to pinpoint the propagation of differences. Therefore, we run the comparison in both condition orders, 
and we label a process as non-reproducible (red) if it creates different results in at least one condition order.}

\subsection{Implementation}
\revised{
\toolname is implemented in Python (>=3.6). In this work we used \toolname version 0.2 and the following version of
the Python package dependencies: NumPy v1.19.0~\cite{oliphant2006guide} and Pandas v1.0.5~\cite{mckinney2011pandas},
for data manipulations, SciPy v1.5.1~\cite{2020SciPy-NMeth} and Scikit-learn v.0.23.1~\cite{pedregosa2011scikit}
for the clustering of provenance graphs, Zss v1.2.0~\cite{zhang1989simple} for tree distances,
\reprozip v1.0.11 for the capture of provenance traces, Docker v17.05~\cite{merkel2014docker} for
the edition of container images, and Boutiques v0.5.25~\cite{glatard2018boutiques} for uniform pipeline executions.

Software users will mostly have to interact with the Boutiques and \reprozip packages.
Boutiques is a flexible description framework for containerized pipelines, required by
the pipelines analyzed in \toolname. It provides a JSON schema to describe inputs, outputs and their dependencies.
Examples, tutorials and usage documentation are available at \url{http://boutiques.github.io}. \reprozip intercepts
system calls to identify the files and processes involved in a pipeline execution.
Before using \toolname, users have to collect \reprozip traces of their pipeline executions.
Examples in the \toolname documentation include \reprozip provenance capture.
More documentation on \reprozip is available at \url{https://www.reprozip.org}.}

@@

We randomly selected 20 unprocessed subjects from the HCP data release S500
available in \href{https://db.humanconnectome.org}{the ConnectomDB
repository} as a subset of the 1200 Subject Release \revised{(see Supplementary Table S1)}.
For each subject, available data consisted of 1 or 2 T1-weighted images and 1 or 2
T2-weighted images, with $256 \times 320 \times 320$ voxels of size $0.7
\times 0.7 \times 0.7$ mm. Acquisition protocols and parameters are
detailed in~\cite{van2013wu}.

@@

We processed the 20 subjects with PreFreeSurfer and FreeSurfer, using the 2 CentOS versions.
\revised{The PreFreesurfer results obtained in CentOS6 were used as the input of FreeSurfer in both conditions. 
We also used the \reprozip trace file captured in CentOS6 for labeling the processes in both pipelines.}
Each subject was processed twice on the same operating
system to detect within-OS variability coming from pseudo-random
operations. We compared pipeline results using  FreeSurfer tools \texttt{mri\_diff},
\texttt{mris\_diff}, and \texttt{lta\_diff}, to ignore execution-specific information such as file path or
timestamps. To compare segmentations $X$ and $Y$, we used the Dice coefficient defined as follows:
\[DICE=\frac{2|X \cap Y|}{|X| + |Y|}\]

\revised{
The Dice coefficient~\cite{dice1945measures} is a commonly used metric to validate medical image segmentation.
Dice values range from 0 to 1, with 1 indicating a perfect overlap between two segmentation results and 0 indicating no overlap.
Alternatively, the Jaccard coefficient~\cite{jaccard1912distribution} could be used;
there is a direct correspondence between both metrics.}

@@

\section{Results}
\revised{
All experiments were run on a machine with a 3.4GHz, 8-core  Intel Core i7 processor, 32GB of RAM, 
CentOS 7.3.1611, and Linux kernel version 3.10.
The processing time, output file size, number of file accesses and number of processes observed in PreFreeSurfer
and FreeSurfer are shown in~Table~\ref{table:pipeline-stats}.
The scripts and analyses used to create the figures in this section are available at
\url{https://github.com/big-data-lab-team/HCP-reproducibility-paper}.

@@

                          & \makecell{Mean}  &  \makecell{Standard error}  & \makecell{Mean}   &  \makecell{Standard error}   \\ \midrule
  Processing time (mins)  &     106.67       &     2.68                    &     650.25        &     8.88     \\
  Output file size (GB)   &     2.8          &     0.10                    &     4.15          &     0.15     \\
  Number of file accesses &     94,089       &     2,645                   &     62,729        &     984     \\
  Number of processes     &     8,731         &     198                     &     4,031          &     47     \\
  \bottomrule
  \end{tabular}
  \end{threeparttable}
  \end{table*}
}
% All experiments were run on a machine with a 3.4GHz Intel Core i7 processor
% and 32GB memory under the CentOS7 operating system.
% The average processing time per subject was approximately 2~hours (with standard error X)
% for PreFreeSurfer and 8~hours (with standard error X) for FreeSurfer.
% The average output file size was 2.7~GB (with standard error Y) for PreFreeSurfer
% and 4.1~GB (with standard error Y) for FreeSurfer.
%\tristan{Ali, are these numbers per subject or for 20 subjects? it's an average of 20 subjects}
% For each subject, PreFreeSurfer accessed 83K files and created 7.7K processes, and FreeSurfer accessed 62K files and
% created 4K processes.

@@

Figure~\ref{fig:pfs_heatmap} shows the frequency of non-reproducible pipeline processes
in PreFreeSurfer.
The processes identified as non-reproducible were observed in linear registration
with FSL \flirt (in ACPC-Alignment, Brain Extraction, Distortion Correction, and
Atlas Registration), in non-linear registration with FSL \fnirt (in Brain Extraction
and Atlas Registration), and in image warping with FSL \texttt{new\_invwarp} (in Brain Extraction
and Atlas Registration). Differences were also observed in image mean
computations with FSL \texttt{maths}  (in Anatomical Average).
Figure~\ref{fig:complete_pfs} shows a complete PreFreeSurfer labeled DAG, localizing
the observed differences in the entire pipeline, for a given subject.
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{./figures/heatmap.pdf}
  \caption{Heatmap of non-reproducible processes across PreFreeSurfer pipeline steps.
  Each cell represents the occurrence of a particular command line in a
  pipeline step among Anatomical Average (AAve), Anterior/Posterior
  Commissure Alignment (ACPC-A), Brain Extraction (BExt), Bias Field
  Correction (BFC), or Atlas-Registration (AR). Cell labels indicate the
  fraction of subjects for which the corresponding process wasn't reproducible. For example,
  the \flirt tool was invoked 6 times in step DC for each of the 20
  subjects: 2 instances weren't reproducible in 19 subjects, 3
  instances were always reproducible, and 1 instance wasn't reproducible in
  17 subjects.
  \revised{Grey cells indicate that the process did not occur in the corresponding pipeline step.}}
  \label{fig:pfs_heatmap}
\end{figure*}

\begin{figure*}
   \centering
    \includegraphics[width=\linewidth]{figures/pfs-labeled.pdf}
    \caption{A complete provenance graph from the PreFreesurfer pipeline.
    \revised{Node labels use the same abbreviations as in Figure~\ref{fig:pfs_heatmap}.}
    For better visualization, processes associated with commands
    in \texttt{/bin} or \texttt{/usr/bin} were omitted, as well as
  \texttt{imtest}, \texttt{imcp}, \texttt{remove\_ext}, \texttt{fslval}, \texttt{avscale}, and \texttt{fslhd}.}
    \label{fig:complete_pfs}
\end{figure*}

\revised{
Figure~\ref{fig:fnirt_result} compares \fnirt results in Brain Extraction
for a particular subject using the checkerboard pattern,
a common method to illustrate the magnitude of the differences in registration results.
Differences appear to be visually important, in particular in the areas framed in red, 
to the point that most experimenters would likely reject such a registration following visual quality control.}

\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{figures/t2w_alignment.png}
    \caption{Differences between T2 \fnirt results in PreFreeSurfer's Brain Extraction (CentOS6 vs CentOS7).
    \revised{
    The colored squares indicate results obtained with CentOS6 (in purple) and CentOS7 (in green).
    The red boxes highlight regions with significant differences between the two OSes. }
    An animated version of the comparison is available \href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/pfs_t2w_alignment.gif}{here} for better visualization.}
    \label{fig:fnirt_result}
\end{figure}

@@

However, FreeSurfer results still differ between conditions, due to the
propagation of differences created in PreFreeSurfer. We
observed the effect of this propagation in FreeSurfer results, as shown in
Figure~\ref{fig:tissue_class} for whole-brain segmentations. The
Dice coefficients associated with the 44 regions segmented by
FreeSurfer are shown in Figure~\ref{fig:scatter_plot}, showing that Dice coefficients below 0.9 are
observed in most regions, and particularly in the smallest ones.
\revised{However, no significant correlation between the Dice values and the region sizes was found
(Pearson's coefficient = 0.12, p-value = 0.43).}

\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{figures/brain_segmentation_mni.png}
  \caption{Sum of binarized differences between whole-brain FreeSurfer
  segmentations obtained from PreFreeSurfer processings in CentOS6 vs CentOS7
   (N=20). Segmentations were resampled and overlaid to the MNI152 volume
  template.
  \revised{
  Each voxel shows the number of subjects for which different results were observed between CentOS 6 and CentOS 7.}
  An animated comparison of segmentations obtained for a particular subject is available
\href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/fs_brain_segmentation.gif}
{here} for better visualization.}
  \label{fig:tissue_class}
\end{figure}

\begin{figure*}
  \hspace*{-1cm}
  \includegraphics[width=1.1\linewidth]{figures/dice_regions.pdf}
    \caption{Dice coefficients between regions segmented by FreeSurfer in CentOS6 vs CentOS7 (N=20), ordered by increasing
    median values.
    \revised{Each point represents the Dice coefficient between segmentations of a particular region obtained in CentOS 6 vs CentOS 7 for a given subject.}
    Boxes brightness is proportional to the logarithm of the corresponding brain region size.}
    \label{fig:scatter_plot}
  \end{figure*}

@@

\revised{
To evaluate the reproducibility of a pipeline, \toolname needs to execute it 5 times 
in order to (1) record a first \reprozip trace, (2) save transient files in the first condition, 
(3) compare results in the second condition, and repeat steps (2) and (3) for the other order of execution. 
It might be possible to further reduce this overhead by executing at step (2) only the processes 
depending on transient files, and capturing the transient files for the second condition simultaneously at step (3).}

We demonstrated the applicability of our approach by evaluating two of the
arguably most complex pipelines in neuroimaging. Technically, these
pipelines consist of a mix of tools assembled from different toolboxes
through a variety of scripts written in different languages. Our file-based
approach, notably enabled by \reprozip, was able to analyze these pipelines
without requiring their instrumentation, which saved a very substantial
technical effort. The assumptions made on the pipeline structure, related
to the absence of concurrent writes, were not violated in our analysis, and
are likely to not impede \toolname's applicability to the most common
neuroimaging pipelines.

\revised{
\toolname only tests pipeline reproducibility in the scope of a particular dataset. However, it is very plausible 
for pipeline processes to exhibit different reproducibility behaviors when executed on different datasets. 
Therefore, only the lack of reproducibility of a pipeline process could be guaranteed from an analysis with \toolname, 
since proving reproducibility would require testing the pipeline on all possible datasets, in all possible environments, 
which is not feasible. Two elements could be considered in future work to address this issue. 
First, similar to conventional software testing, a code coverage metric could be developed to assess 
the fraction of the pipeline code involved in the tested dataset and parameters. This would quantify 
the representativity of the dataset and pipeline parameters used in the evaluation. Second, statistical 
risk models could be used to estimate the probability for a process to be reproducible, given a set of 
observations with no numerical differences. For instance, models described in~\cite{beauzamy2004methodes} 
could be leveraged for this purpose.}

@@

\revised{
Computational environments are only one of many factors contributing to the
on-going reproducibility crisis. In fact, sample size selection,
publication bias, or methodological flexibility in the analysis are likely
to have a stronger effect than numerical perturbations, although
to our knowledge no evidence of this is available. We refer to the
studies in~\cite{botvinik2020variability,bowring2019exploring,bhagwat2020understanding,kennedy2019everything} for deeper
analyses of the associated effects on neuroimaging analyses. It should also be
noted that the effects of computational environments and these other
factors manifest at different levels: referring to the terminology used
in~\cite{peng2011reproducible}, computational environments are associated with
\emph{reproducibility}, the minimal standard by which identical results
should be obtainable from identical data and parameters, while the other
aforementioned factors belong to \emph{replicability}, the ultimate
standard by which independent experimenters should be able to draw similar
conclusions from similar experiments. In practice, variability resulting
from computational environments manifests during software testing (test
results depend on execution platform), deployment on HPC systems (results
obtained on local vs HPC systems differ), or software version updates
(results obtained before vs after the update differ), while factors related
to replicability impact the community more broadly. Ultimately, both
reproducibility and replicability should be understood and improved.
}

@@

\revised{
\section{Availability of Source Code and Requirements}
\begin{itemize}
  \item Project name: \toolname
  \item Project home page: \url{https://github.com/big-data-lab-team/spot}
  \item Operating system: Linux
  \item Programming language: Python (3.6 or higher)
  \item Main dependencies: \reprozip, Docker, and Boutiques
  \item Other dependencies: see \href{https://github.com/big-data-lab-team/spot}{\texttt{setup.py}}
  \item License: MIT License
  \item Biotools identifier: \href{https://bio.tools/spottool}{spottool}
  \item SciCrunch ID: \href{https://scicrunch.org/scicrunch/Resources/record/nlx_144509-1/SCR_018915/resolver}{RRID:SCR\_018915}
  \item DOI: \href{https://doi.org/10.5281/zenodo.3873219}{10.5281/zenodo.3873219}

@@

\section{Supplementary Data}
Supplementary materials include Supplementary Table S1 that contains a summary of the subjects used in the experiments.
}

\section{Acknowledgments}
\revised{We thank the reviewers for their thorough and insightul review which greatly contributed to the improvement 
of the paper.}
We warmly thank Compute Canada (\url{http://www.computecanada.ca}) and Calcul
Qu\'ebec (\url{http://www.calculquebec.ca}) for providing the infrastructure used in our experiments.
Data and pipelines were provided by the Human Connectome Project, WU-Minn
Consortium (Principal Investigators: David Van Essen and Kamil Ugurbil;
1U54MH091657) funded by the 16 NIH Institutes and Centers that support
the NIH Blueprint for Neuroscience Research; and by the McDonnell
Center for Systems Neuroscience at Washington University.
%\tristan{check ethics}

>>>>

%%%% Commands %%%%
% \newcommand{\todo}[1]{\color{red}\textbf{TODO:}#1\color{black}}
% \newcommand{\note}[2]{\color{blue}Note from \textsc{#1}: #2\color{black}}
% \newcommand{\revised}[1]{\color{blue}#1\color{black}\xspace}
\newcommand{\reprozip}[0]{ReproZip\xspace}
\newcommand{\tristan}[1]{\color{orange}\textbf{From Tristan:}#1\color{black}}
\newcommand{\toolname}[0]{Spot\xspace}
\newcommand{\flirt}[0]{\texttt{flirt}\xspace}
\newcommand{\fnirt}[0]{\texttt{fnirt}\xspace}

@@

Numerical perturbations resulting from variations in computational
environments impact data analyses in various fields, but identifying the
origin of these perturbations in complex pipelines remains challenging.  In
some cases, small perturbations resulting from changes in operating system
versions~\cite{Glatard2015}, hardware~\cite{jezequel2015estimation}, or
parallelization parameters~\cite{diethelm2011limits}, result in
substantially different analysis outcomes, due to the propagation
and amplification of floating-point errors. While the existence of such
numerical errors is well known~\cite{stoer2013introduction}, their impact
on scientific computations has multiplied with the rise of the Big Data
era, due to the sustained growth of data sets, the increasing complexity of
analysis pipelines, and the diversification of computing infrastructures.
To better understand and
correct these effects, efficient tools are
needed to assist pipeline developers
in the comparison of results obtained across different conditions.

@@

\begin{listing}
  \inputminted{bash}{"data/example/example.sh"}
  \caption{Example pipeline that computes the volume of the brain from a T1 image.}
  \label{listing:sample-script}
\end{listing}

@@

We use \reprozip~\cite{rampin2016reprozip}
to capture: (1) the set of processes created by the
pipeline, and
(2) the set of files read and written by each process, including
temporary files. \reprozip collects this information through the
\texttt{ptrace()} system call, with no required instrumentation of the pipeline.
Using the \reprozip trace, \toolname reconstructs a provenance graph by creating process and file
nodes and by adding directed edges corresponding
to file reads and writes (Figure~\ref{fig:provenance-graph}).
We assume that provenance graphs are identical for the \reprozip traces obtained from
the same subjects in different operating systems.

@@

The labeled graph can differ depending on the order of executions in which condition we capture transient files 
or execute the pipeline to pinpoint the propagation of differences. Therefore, we run the comparison in both condition orders, 
and we label a process as non-reproducible (red) if it creates different results in at least one condition order.

\subsection{Implementation}
\toolname is implemented in Python (>=3.6). In this work we used \toolname version 0.2 and the following version of
the Python package dependencies: NumPy v1.19.0~\cite{oliphant2006guide} and Pandas v1.0.5~\cite{mckinney2011pandas},
for data manipulations, SciPy v1.5.1~\cite{2020SciPy-NMeth} and Scikit-learn v.0.23.1~\cite{pedregosa2011scikit}
for the clustering of provenance graphs, Zss v1.2.0~\cite{zhang1989simple} for tree distances,
\reprozip v1.0.11 for the capture of provenance traces, Docker v17.05~\cite{merkel2014docker} for
the edition of container images, and Boutiques v0.5.25~\cite{glatard2018boutiques} for uniform pipeline executions.

Software users will mostly have to interact with the Boutiques and \reprozip packages.
Boutiques is a flexible description framework for containerized pipelines, required by
the pipelines analyzed in \toolname. It provides a JSON schema to describe inputs, outputs and their dependencies.
Examples, tutorials and usage documentation are available at \url{http://boutiques.github.io}. \reprozip intercepts
system calls to identify the files and processes involved in a pipeline execution.
Before using \toolname, users have to collect \reprozip traces of their pipeline executions.
Examples in the \toolname documentation include \reprozip provenance capture.
More documentation on \reprozip is available at \url{https://www.reprozip.org}.

@@

We randomly selected 20 unprocessed subjects from the HCP data release S500
available in \href{https://db.humanconnectome.org}{the ConnectomDB
repository} as a subset of the 1200 Subject Release (see Supplementary Table S1).
For each subject, available data consisted of 1 or 2 T1-weighted images and 1 or 2
T2-weighted images, with $256 \times 320 \times 320$ voxels of size $0.7
\times 0.7 \times 0.7$ mm. Acquisition protocols and parameters are
detailed in~\cite{van2013wu}.

@@

We processed the 20 subjects with PreFreeSurfer and FreeSurfer, using the 2 CentOS versions.
The PreFreesurfer results obtained in CentOS6 were used as the input of FreeSurfer in both conditions. 
We also used the \reprozip trace file captured in CentOS6 for labeling the processes in both pipelines.
Each subject was processed twice on the same operating
system to detect within-OS variability coming from pseudo-random
operations. We compared pipeline results using  FreeSurfer tools \texttt{mri\_diff},
\texttt{mris\_diff}, and \texttt{lta\_diff}, to ignore execution-specific information such as file path or
timestamps. To compare segmentations $X$ and $Y$, we used the Dice coefficient defined as follows:
\[DICE=\frac{2|X \cap Y|}{|X| + |Y|}\]

The Dice coefficient~\cite{dice1945measures} is a commonly used metric to validate medical image segmentation.
Dice values range from 0 to 1, with 1 indicating a perfect overlap between two segmentation results and 0 indicating no overlap.
Alternatively, the Jaccard coefficient~\cite{jaccard1912distribution} could be used;
there is a direct correspondence between both metrics.

@@

\section{Results}
All experiments were run on a machine with a 3.4GHz, 8-core  Intel Core i7 processor, 32GB of RAM, 
CentOS 7.3.1611, and Linux kernel version 3.10.
The processing time, output file size, number of file accesses and number of processes observed in PreFreeSurfer
and FreeSurfer are shown in~Table~\ref{table:pipeline-stats}.
The scripts and analyses used to create the figures in this section are available at
\url{https://github.com/big-data-lab-team/HCP-reproducibility-paper}.

@@

                          & \makecell{Mean}  &  \makecell{Standard error}  & \makecell{Mean}   &  \makecell{Standard error}   \\ \midrule
  Processing time (mins)  &     106.67       &     2.68                    &     650.25        &     8.88     \\
  Output file size (GB)   &     2.8          &     0.10                    &     4.15          &     0.15     \\
  Number of file accesses &     94,089       &     2,645                   &     62,729        &     984     \\
  Number of processes     &     8,731         &     198                     &     4,031          &     47     \\
  \bottomrule
  \end{tabular}
  \end{threeparttable}
  \end{table*}

% All experiments were run on a machine with a 3.4GHz Intel Core i7 processor
% and 32GB memory under the CentOS7 operating system.
% The average processing time per subject was approximately 2~hours (with standard error X)
% for PreFreeSurfer and 8~hours (with standard error X) for FreeSurfer.
% The average output file size was 2.7~GB (with standard error Y) for PreFreeSurfer
% and 4.1~GB (with standard error Y) for FreeSurfer.
%\tristan{Ali, are these numbers per subject or for 20 subjects? it's an average of 20 subjects}
% For each subject, PreFreeSurfer accessed 83K files and created 7.7K processes, and FreeSurfer accessed 62K files and
% created 4K processes.

@@

Figure~\ref{fig:pfs_heatmap} shows the frequency of non-reproducible pipeline processes
in PreFreeSurfer.
The processes identified as non-reproducible were observed in linear registration
with FSL \flirt (in ACPC-Alignment, Brain Extraction, Distortion Correction, and
Atlas Registration), in non-linear registration with FSL \fnirt (in Brain Extraction
and Atlas Registration), and in image warping with FSL \texttt{new\_invwarp} (in Brain Extraction
and Atlas Registration). Differences were also observed in image mean
computations with FSL \texttt{maths}  (in Anatomical Average).
Figure~\ref{fig:complete_pfs} shows a complete PreFreeSurfer labeled DAG, localizing
the observed differences in the entire pipeline, for a given subject.
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{./figures/heatmap.pdf}
  \caption{Heatmap of non-reproducible processes across PreFreeSurfer pipeline steps.
  Each cell represents the occurrence of a particular command line in a
  pipeline step among Anatomical Average (AAve), Anterior/Posterior
  Commissure Alignment (ACPC-A), Brain Extraction (BExt), Bias Field
  Correction (BFC), or Atlas-Registration (AR). Cell labels indicate the
  fraction of subjects for which the corresponding process wasn't reproducible. For example,
  the \flirt tool was invoked 6 times in step DC for each of the 20
  subjects: 2 instances weren't reproducible in 19 subjects, 3
  instances were always reproducible, and 1 instance wasn't reproducible in
  17 subjects.
  Grey cells indicate that the process did not occur in the corresponding pipeline step.}
  \label{fig:pfs_heatmap}
\end{figure*}

\begin{figure*}
   \centering
    \includegraphics[width=\linewidth]{figures/pfs-labeled.pdf}
    \caption{A complete provenance graph from the PreFreesurfer pipeline.
    Node labels use the same abbreviations as in Figure~\ref{fig:pfs_heatmap}.
    For better visualization, processes associated with commands
    in \texttt{/bin} or \texttt{/usr/bin} were omitted, as well as
  \texttt{imtest}, \texttt{imcp}, \texttt{remove\_ext}, \texttt{fslval}, \texttt{avscale}, and \texttt{fslhd}.}
    \label{fig:complete_pfs}
\end{figure*}

Figure~\ref{fig:fnirt_result} compares \fnirt results in Brain Extraction
for a particular subject using the checkerboard pattern,
a common method to illustrate the magnitude of the differences in registration results.
Differences appear to be visually important, in particular in the areas framed in red, 
to the point that most experimenters would likely reject such a registration following visual quality control.

\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{figures/t2w_alignment.png}
    \caption{Differences between T2 \fnirt results in PreFreeSurfer's Brain Extraction (CentOS6 vs CentOS7).
    The colored squares indicate results obtained with CentOS6 (in purple) and CentOS7 (in green).
    The red boxes highlight regions with significant differences between the two OSes. 
    An animated version of the comparison is available \href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/pfs_t2w_alignment.gif}{here} for better visualization.}
    \label{fig:fnirt_result}
\end{figure}

@@

However, FreeSurfer results still differ between conditions, due to the
propagation of differences created in PreFreeSurfer. We
observed the effect of this propagation in FreeSurfer results, as shown in
Figure~\ref{fig:tissue_class} for whole-brain segmentations. The
Dice coefficients associated with the 44 regions segmented by
FreeSurfer are shown in Figure~\ref{fig:scatter_plot}, showing that Dice coefficients below 0.9 are
observed in most regions, and particularly in the smallest ones.
However, no significant correlation between the Dice values and the region sizes was found
(Pearson's coefficient = 0.12, p-value = 0.43).

\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{figures/brain_segmentation_mni.png}
  \caption{Sum of binarized differences between whole-brain FreeSurfer
  segmentations obtained from PreFreeSurfer processings in CentOS6 vs CentOS7
   (N=20). Segmentations were resampled and overlaid to the MNI152 volume
  template.
  Each voxel shows the number of subjects for which different results were observed between CentOS 6 and CentOS 7.
  An animated comparison of segmentations obtained for a particular subject is available
\href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/fs_brain_segmentation.gif}
{here} for better visualization.}
  \label{fig:tissue_class}
\end{figure}

\begin{figure*}
  \hspace*{-1cm}
  \includegraphics[width=1.1\linewidth]{figures/dice_regions.pdf}
    \caption{Dice coefficients between regions segmented by FreeSurfer in CentOS6 vs CentOS7 (N=20), ordered by increasing
    median values.
    Each point represents the Dice coefficient between segmentations of a particular region obtained in CentOS 6 vs CentOS 7 for a given subject.
    Boxes brightness is proportional to the logarithm of the corresponding brain region size.}
    \label{fig:scatter_plot}
  \end{figure*}

@@

To evaluate the reproducibility of a pipeline, \toolname needs to execute it 5 times 
in order to (1) record a first \reprozip trace, (2) save transient files in the first condition, 
(3) compare results in the second condition, and repeat steps (2) and (3) for the other order of execution. 
It might be possible to further reduce this overhead by executing at step (2) only the processes 
depending on transient files, and capturing the transient files for the second condition simultaneously at step (3).

The target users of the \toolname tool are primarily pipeline developers and users who have 
technical skills for creating Docker containers and Boutiques JSON files.
We demonstrated the applicability of our approach by evaluating two of the
arguably most complex pipelines in neuroimaging. Technically, these
pipelines consist of a mix of tools assembled from different toolboxes
through a variety of scripts written in different languages. Our file-based
approach, notably enabled by \reprozip, was able to analyze these pipelines
without requiring their instrumentation, which saved a very substantial
technical effort. The assumptions made on the pipeline structure, related
to the absence of concurrent writes, were not violated in our analysis, and
are likely to not impede \toolname's applicability to the most common
neuroimaging pipelines.

\toolname only tests pipeline reproducibility in the scope of a particular dataset. However, it is very plausible 
for pipeline processes to exhibit different reproducibility behaviors when executed on different datasets. 
Therefore, only the lack of reproducibility of a pipeline process could be guaranteed from an analysis with \toolname, 
since proving reproducibility would require testing the pipeline on all possible datasets, in all possible environments, 
which is not feasible. Two elements could be considered in future work to address this issue. 
First, similar to conventional software testing, a code coverage metric could be developed to assess 
the fraction of the pipeline code involved in the tested dataset and parameters. This would quantify 
the representativity of the dataset and pipeline parameters used in the evaluation. Second, statistical 
risk models could be used to estimate the probability for a process to be reproducible, given a set of 
observations with no numerical differences. For instance, models described in~\cite{beauzamy2004methodes} 
could be leveraged for this purpose.

@@

Computational environments are only one of many factors contributing to the
on-going reproducibility crisis. In fact, sample size selection,
publication bias, or methodological flexibility in the analysis are likely
to have a stronger effect than numerical perturbations, although
to our knowledge no evidence of this is available. We refer to the
studies in~\cite{botvinik2020variability,bowring2019exploring,bhagwat2020understanding,kennedy2019everything} for deeper
analyses of the associated effects on neuroimaging analyses. It should also be
noted that the effects of computational environments and these other
factors manifest at different levels: referring to the terminology used
in~\cite{peng2011reproducible}, computational environments are associated with
\emph{reproducibility}, the minimal standard by which identical results
should be obtainable from identical data and parameters, while the other
aforementioned factors belong to \emph{replicability}, the ultimate
standard by which independent experimenters should be able to draw similar
conclusions from similar experiments. In practice, variability resulting
from computational environments manifests during software testing (test
results depend on execution platform), deployment on HPC systems (results
obtained on local vs HPC systems differ), or software version updates
(results obtained before vs after the update differ), while factors related
to replicability impact the community more broadly. Ultimately, both
reproducibility and replicability should be understood and improved.

@@

\section{Availability of Source Code and Requirements}
\begin{itemize}
  \item Project name: \toolname
  \item Project home page: \url{https://github.com/big-data-lab-team/spot}
  \item Operating system: Linux
  \item Programming language: Python (3.6 or higher)
  \item Main dependencies: \reprozip, Docker, and Boutiques
  \item Other dependencies: see \href{https://github.com/big-data-lab-team/spot}{\texttt{setup.py}}
  \item License: MIT License
  \item Biotools identifier: \href{https://bio.tools/spottool}{spottool}
  \item SciCrunch ID: \href{https://scicrunch.org/scicrunch/Resources/record/nlx_144509-1/SCR_018915/resolver}{RRID:SCR\_018915}
  \item DOI: \href{https://doi.org/10.5281/zenodo.3873219}{10.5281/zenodo.3873219}

@@

\section{Availability of Supporting Data}
Snapshots of our code and other supporting data are openly available in the GigaScience repository, GigaDB~\cite{salari2020global}.

\section{Supplementary Data}
Supplementary materials include Supplementary Table S1 that contains a summary of the subjects used in the experiments.

\section{Acknowledgments}
We thank the reviewers for their thorough and insightul review which greatly contributed to the improvement 
of the paper.
We warmly thank Compute Canada (\url{http://www.computecanada.ca}) and Calcul
Qu\'ebec (\url{http://www.calculquebec.ca}) for providing the infrastructure used in our experiments.
Data and pipelines were provided by the Human Connectome Project, WU-Minn
Consortium (Principal Investigators: David Van Essen and Kamil Ugurbil;
1U54MH091657) funded by the 16 NIH Institutes and Centers that support
the NIH Blueprint for Neuroscience Research; and by the McDonnell
Center for Systems Neuroscience at Washington University.
%\tristan{check ethics}
